---
title: "Organisations"
vignette: >
  %\VignetteIndexEntry{Organisations}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Why bother setting up an `org_list`?

`citeme` uses the `org_list` to check whether a project complies with the
organisation's rules about naming conventions, copyright holders, funders,
and publishers.
This allows an organisation to enforce a consistent set of rules across all
its projects.

## The organisation-level `citeme` repository

`citeme` allows every organisation to publish its rules in a single public git repository named `citeme`, located directly under the organisation's git host account.
For an account at https://gitlab.com/ThierryO, it is https://gitlab.com/ThierryO/citeme.
We created https://gitlab.com/ThierryO/citeme as a toy example to illustrate the structure of such a repository.
https://github.com/inbo/citeme is not an example because it hosts the `citeme` package itself, which hard-codes the INBO rules in `inbo_org_list()`.

The repository holds the organisation's settings, not code:

| File or folder | Purpose |
|---|---|
| `organisation.yml` | The `org_list`: names, emails, ROR, roles and allowed licences. |
| `set_organisation.R` | The optional script that generates `organisation.yml`, kept under version control so the settings are reproducible. |
| `pkgdown.css`, `pkgdown/` | Optional house style and images for `pkgdown` websites. |

The repository must be public and must have a `main` branch.
`citeme` clones it anonymously, so a private repository cannot be read by
your colleagues' sessions.

### Benefits

- **One source of truth.**
  Every project in the organisation reads the same `organisation.yml`, so names, email addresses, ROR identifiers and licence lists cannot drift apart between repositories.
- **Rules travel automatically.**
  A colleague runs `get_default_org_list()` once and gets the organisation's conventions, without reading any policy document.
- **Change once, apply everywhere.**
  Update `organisation.yml` in the `citeme` repository and every project picks up the new rules the next time its cache is refreshed.
- **Enforceable in CI.**
  `org_list$check()` recognises a continuous integration environment, so a non-compliant copyright holder, funder, publisher or licence can block a
  pull request or a release.
- **Licences stay legal.**
  The allowed licences are listed as URLs to the licence text, so each
  project gets a byte-identical `LICENSE.md`.
- **Reproducible settings.**
  Keeping `set_organisation.R` next to the generated `organisation.yml`
  documents how the rules came about and makes review of a change easy.
- **Reusable for other purposes.**
  The organisation's settings can be reused in other projects or scripts, ensuring consistency across different tools and workflows.
  The `citeme` repository can be used to store additional resources, such as templates, guidelines, or scripts that support the organisation's practices.
  For instance the `checklist` package uses the `citeme` repository to store its mandatory configuration files at the organisation level.

### Setting up the repository

1. Create a new **public** repository named `citeme` directly under your organisation's git account, with `main` as the default branch.
2. Write a script, for example `set_organisation.R`, that builds the `org_list` for your organisation.
   Create one `org_item` per organisation that may appear as `rightsholder`, `funder` or `publisher`, then combine them with `org_list$new()`.
   Set `git` to the URL of the organisation account, *without* the `citeme` suffix.

```{r}
#| label: setup-org-script
#| eval: false
library(citeme)
org_item$new(
  name = c(
    `en-GB` = "The checklist organisation",
    `nl-BE` = "De checklist organisatie"
  ),
  email = "info@organisation.checklist",
  rightsholder = "shared",
  funder = "shared",
  license = list(
    package = c(
      MIT = paste(
        "https://raw.githubusercontent.com/inbo/citeme/refs/heads/main",
        "inst/licenses/mit.md",
        sep = "/"
      )
    ),
    project = character(0),
    data = character(0)
  )
) |>
  org_list$new(git = "https://gitlab.com/ThierryO") -> org
org$write()
```

3. Run the script from the root of the clone of the `citeme` repository. `write()` creates `organisation.yml` in the working directory.
4. Commit both `set_organisation.R` and `organisation.yml` and push them to `main`.
5. Optionally add the licence texts you refer to (as Markdown files) and a `pkgdown.css` with your house style to the same repository.
6. Verify the result from a project of your organisation:

```{r}
#| label: setup-org-verify
#| eval: false
get_default_org_list(".")
```

This clones the organisation's `citeme` repository and caches it under `tools::R_user_dir("citeme", "config")`.
Each colleague runs this once, and again whenever the organisation's rules change.

Compare your result with the toy example at <https://gitlab.com/ThierryO/citeme>, which contains exactly these pieces: a `set_organisation.R` script, the generated `organisation.yml` and some `pkgdown` styling.

## The `org_item` class

The `org_item` class stores information about a single organisation.
It requires two mandatory fields: `name` and `email`.

- `name` must be a named character vector where each name is a [BCP 47](https://en.wikipedia.org/wiki/IETF_language_tag) language code (e.g. `"en-GB"`, `"nl-BE"`).
  This allows the organisation name to be used in different languages.
  The first entry is used as the default when no matching language is available.
- `email` serves two purposes:
  1. It gives users a contact point for the organisation when individual authors are unavailable.
  2. `citeme` uses the email domain to detect whether a person is affiliated with the organisation.

Optionally you can provide:

- A [ROR identifier](https://ror.org/) (`ror`)
- A website URL (`website`)
- A logo URL (`logo`)
- Whether affiliated persons must have an
  [ORCID identifier](https://orcid.org/) (`orcid`)
- A [Zenodo community](https://zenodo.org/communities#community-info)
  identifier (`zenodo`)
- Allowed licenses per project type (`license`)

### Roles: `rightsholder`, `funder`, and `publisher`

`org_item` defines the required status of the organisation for each of the three roles: `rightsholder`, `funder`, and `publisher`.
Each accepts one of the following values:

| Value | Meaning |
|---|---|
| `"optional"` | The organisation may or may not be listed in this role. |
| `"single"` | The organisation must be listed and no other organisation may be listed. |
| `"shared"` | The organisation must be listed; other organisations may be listed too. |
| `"when no other"` | The organisation must be listed only if no other organisation is listed. |

The default for all three roles is `"optional"`.

## The `org_list` class

The `org_list` class holds a collection of `org_item` objects.
The rules of the different items must be mutually compatible.
For example, if one organisation has `rightsholder = "single"`, no other organisation can be set to `"single"` or `"shared"` for the same role.

The `org_list` also stores the URL of the organisation's git host.
`citeme` looks for an `organisation.yml` file in the `citeme` repository at that URL to retrieve the minimum rules for the organisation.
For example, when `git = "https://github.com/inbo"`, `citeme` looks at `https://github.com/inbo/citeme`.

## Creating an `org_list` programmatically

Use `org_item$new()` to create `org_item` objects.
Note that license values must be URLs pointing to publicly accessible Markdown files with the license text.
We recommend hosting these files in your organisation's `citeme` repository.
`citeme` caches license files locally and includes them as `LICENSE.md` in packages and projects.

```{r}
#| label: create-org-item
library(citeme)

inbo <- org_item$new(
  name = c(
    "nl-BE" = "Instituut voor Natuur- en Bosonderzoek (INBO)",
    "en-GB" = "Research Institute for Nature and Forest (INBO)"
  ),
  email = "info@inbo.be",
  website = "https://www.vlaanderen.be/inbo/en-gb",
  logo = "https://inbo.github.io/citeme/reference/figures/logo-en.png",
  ror = "00j54wy13",
  orcid = TRUE,
  zenodo = "inbo",
  rightsholder = "shared",
  funder = "when no other",
  publisher = "when no other",
  license = list(
    package = c(
      `GPL-3.0` = paste(
        "https://raw.githubusercontent.com/inbo/citeme/refs/heads/main",
        "inst/licenses/gplv3.md",
        sep = "/"
      ),
      MIT = paste(
        "https://raw.githubusercontent.com/inbo/citeme/refs/heads/main",
        "inst/licenses/mit.md",
        sep = "/"
      )
    ),
    project = c(
      `CC BY 4.0` = paste(
        "https://raw.githubusercontent.com/inbo/citeme/refs/heads/main",
        "inst/licenses/cc_by_4_0.md",
        sep = "/"
      )
    ),
    data = c(
      `CC0` = paste(
        "https://raw.githubusercontent.com/inbo/citeme/refs/heads/main",
        "inst/licenses/cc0.md",
        sep = "/"
      )
    )
  )
)
inbo

anb <- org_item$new(
  name = c(
    `nl-BE` = "Agentschap voor Natuur en Bos (ANB)",
    `en-GB` = "Agency for Nature & Forests (ANB)"
  ),
  email = "natuurenbos@vlaanderen.be",
  ror = "04wcznf70",
  license = list(
    package = character(0),
    project = character(0),
    data = character(0)
  )
)
anb
```

Then combine the items into an `org_list`:

```{r}
#| label: create-org-list
inbo_org <- org_list$new(inbo, anb, git = "https://github.com/inbo")
inbo_org
```

The final step is to store the `org_list` in a `citeme` repository.
`write()` creates an `organisation.yml` file in the given directory.
Commit and push this file to the remote repository.

```{r}
#| label: store-org-list
path_to_citeme_repo <- tempfile("citeme")
dir.create(path_to_citeme_repo, recursive = TRUE)
inbo_org$write(path_to_citeme_repo)
```

## Creating an `org_list` interactively

Use `new_org_list()` to build an `org_list` step by step in an interactive session.
`new_org_item()` similarly walks you through creating a single `org_item`.
Both functions are intended for use in the console, not inside scripts.

## Getting the default `org_list` for a project

Run `get_default_org_list()` inside a git project to fetch the default organisation list from the organisation's `citeme` repository.
The function uses the `origin` remote of the project to determine which organisation to look up.

`get_default_org_list()` stores the result in `citeme`'s user-level configuration.
You only need to run it once per organisation, or when the organisation's defaults change.
After that, `org_list$new()$read()` returns the cached information.

```{r}
#| label: get-org-list
#| eval: false
# Set up a demo git repository
test_dir <- tempfile("test")
dir.create(test_dir)
library(gert)
git_init(test_dir)
git_remote_add(
  url = "https://github.com/inbo/test",
  name = "origin",
  repo = test_dir
)
# Fetch the default organisation list from https://github.com/inbo/citeme
my_org <- get_default_org_list(test_dir)
my_org
```

## Adding an `org_item` to an existing `org_list`

Use the `add_item()` method to extend an existing `org_list`.
Call `write()` afterwards to persist the change to the local `organisation.yml` file.

```{r}
#| label: update-org-list
#| eval: false
# Read the current organisation list from a project directory
my_org <- org_list$new()$read(test_dir)
# Add an organisation to the list
my_org <- my_org$add_item(anb)
my_org
# Persist to disc
my_org$write(test_dir)
```

## The built-in INBO organisation list

`citeme` ships with `inbo_org_list()`, a ready-to-use `org_list` for projects affiliated with the Research Institute for Nature and Forest (INBO) and its partner organisations.

```{r}
#| label: inbo-org-list
inbo_org_list()
```

## FAQ

**Can I use `citeme` without git?**

Yes. Without git, `citeme` only checks rules defined in a local `organisation.yml` file.

**Is a `citeme` repository at the organisation level required?**

No. Without an organisation-level repository, no organisation-wide rules are enforced.

**Do I need to add an `organisation.yml` file to my project?**

No, unless you want to add project-specific rules.
Any project-level rules must still comply with the organisation-level rules.
